D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-04-08 06related_conclusion.tex
.

\paragraph*{Comparison with DiracDec~\cite{diracdec}}
Xu \emph{et al}~\cite{diracdec} define a language and an associate and
commutative rewriting system for Dirac notation, and implement their
language and rewriting system in Mathematica. Our approach follows a
similar pattern. However, there are significant differences in terms
of scope, expressiveness, and efficiency. We review these differences
below.

@@

\paragraph*{Comparison with ZX Calculus}
The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,coecke2011interacting}
is a graphical calculus for quantum states. A main appeal of the ZX
calculus is that its foundations are grounded in categorical quantum
mechanics~\cite{DBLP:conf/lics/AbramskyC04}, a powerful framework for
modeling and reasoning about quantum physics. Another main appeal of
the ZX calculus is that ZX diagrams have an operational interpretation
using graph rewriting. There is a large body of work that studies
properties of these rewriting systems, and in particular completeness
(a set of rules is complete if it can prove all valid identities) and
minimality (a complete set of rules is minimal if removing a rule
leads to incompleteness). Two main proof techniques for completeness
are via termination, or via interpretation. In the first case, one
shows that a rewriting system has unique normal forms and that two
expressions are semantically equivalent iff they have the same normal
form---there exists many relaxations of this result---whereas in the
second case one shows completeness by exhibiting a well-behaved
translation to another system for which completeness holds. Both proof
techniques have been used to prove completeness for multiple settings,
including the Clifford fragment~\cite{backens2014zx}, the
Toffoli+Hadamard fragment~\cite{hadzi15zx}, the Clifford+T
fragment~\cite{jeandel2018complete} and the (qubit) universal
fragment~\cite{HNW18,JeandelPV18beyond}. Subsequent works generalize
completeness to qudits~\cite{Poor2023} for a calculus know as ZXW
which encompasses the ZX and ZW calculi~\cite{zwcalculus}, quantum
circuits~\cite{DBLP:conf/lics/ClementHMPV23} or optical quantum
circuits~\cite{clement_et_al:2022}. We refer the interested reader
to~\cite{vandewetering2020zx} for a historical and technical account
of completeness results up to 2020.

Because they are both grounded in rewriting techniques, there are many
similarities between prior work on the ZX calculus and our own work on
labelled Dirac notation. We highlight some key differences. First,
there exist a number of differences between our language and the
concrete fragments of the ZX-calculus for which completeness has been
studied. For instance, our language features big operators, and
operators for dual spaces. A second difference is that we do not claim
completeness: our main theorem shows a reduction from Labelled Dirac
notation to Dirac notation, for which completeness is only partially
known. From a practical point of view, many of the examples handled by
D-Hammer cannot be formalized in existing ZX-calculus tools. On the
contrary, D-Hammer is not competitive on the examples that can be
verified by ZX-calculus. \gb{give a few examples in terms of size and time}

\paragraph*{Comparison with quantum circuit equivalence checking tools}
A main application of the ZX-calculus is to check equivalence of
quantum circuits; for instance, PyZX~\cite{kissinger2020pyzx} is a
popular Python library for automating equational reasoning in the
ZX-calculus.

There are many other tools for equivalence checking of quantum
circuits. Amy~\cite{amy2018towards} uses the path-sums formalism
formalism to check circuits with thousands of T gates. A series of
works~\cite{AutoQ_pldi_2023,AutoQ2023,AutoQ_popl2025} explores
automata-based approaches to verify quantum circuits at scale. Their
main idea is to encode circuits as a compact, finite tree automata,
often with $\mathit{poly}(n)$ transitions (instead of $2^n$
transitions), where $n$ is the number of qubits of the state. By
providing the semantics of quantum gates over this representation,
they achieve practical verification. Their tool \textsc{autoq} is able
to verify circuits with over 100,00 gates, and was recently extended
to support parametrized verification.

>>>>

\paragraph*{Comparison with DiracDec}
Xu \emph{et al}~\cite{diracdec} define a language and an associate and
commutative rewriting system for Dirac notation, and implement their
language and rewriting system in Mathematica. Our approach follows a
similar pattern. However, there are significant differences in terms
of scope, expressiveness, and efficiency. We review these differences
below.

@@

\paragraph*{Comparison with ZX Calculus}
The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,coecke2011interacting}
is a graphical calculus for quantum states. A main appeal of the ZX
calculus is that its foundations are grounded in categorical quantum
mechanics~\cite{DBLP:conf/lics/AbramskyC04}, a powerful framework for
modeling quantum physics. Another main appeal of the ZX calculus is
that it has a natural operational interpretation based on graph
rewriting. There is a large body of work that defines rewriting
systems for fragments/extensions/variants of the ZX calculus, and
studies their theoretical properties, in particular completeness (a
set of rules is complete if it can prove all valid identities) and
minimality (a complete set of rules is minimal if removing a rule
leads to incompleteness). Two main proof techniques for completeness
are via termination, or via interpretation. In the first case, one
shows that a rewriting system has unique normal forms and that two
expressions are semantically equivalent iff they have the same normal
form---there exists many relaxations of this result---whereas in the
second case one shows completeness by exhibiting a well-behaved
translation to another system for which completeness holds. Both proof
techniques have been used to prove completeness for multiple settings,
including the Clifford fragment~\cite{backens2014zx}, the
Toffoli+Hadamard fragment~\cite{hadzi15zx}, the Clifford+T
fragment~\cite{jeandel2018complete} and the (qubit) universal
fragment~\cite{HNW18,JeandelPV18beyond}. Subsequent works generalize
completeness to qudits~\cite{Poor2023}, quantum
circuits~\cite{DBLP:conf/lics/ClementHMPV23} or optical quantum
circuits~\cite{clement_et_al:2022}. We refer the interested reader
to~\cite{vandewetering2020zx} for a historical and technical account
of completeness results up to 2020.

There are many similarities between the ZX calculus and our formalism
for Labelled Dirac notation. However, there are also some differences
between our formalism and the ones used in ZX calculus. In particular,
our language supports big operators, tensors, and various operations
on Hilbert spaces. These features are typically not considered in
prior work on the ZX calculus, and as a consequence many examples
handled by D-Hammer lack an immediate translation into ZX-calculus.
This additional generality comes with a price. On the one hand, our
theoretical results are weaker: we do not claim completeness or
minimality. Similarly, practical implementations of the ZX-calculus,
such as PyZX~\cite{kissinger2020pyzx}, outperform D-Hammer on examples
that can be handled by both tools, as shown in \Cref{sec:eval}.

\paragraph*{Comparison with other tools}
Beyond ZX calculus, there exists many other tools for simplifying and
proving equivalence of quantum circuits; we refer the reader to recent
surveys~\cite{QPV-survey2021,CharetonLVVBX23} for detailed accounts.
Notable works include~\cite{amy2018towards}, which uses the path-sums
formalism to check circuits with 1,000 of T gates,
and~\cite{AutoQ_pldi_2023,AutoQ2023,AutoQ_popl2025}, which uses
automata-based approaches to verify quantum circuits at scale---their
tool \textsc{autoq} is able to verify circuits with over 100,00 gates,
and was recently extended to support parametrized verification.
